### 2013-02-25 patchy-selection-paper.tex
WorkingWiki importation

Now we can add in the spatial motion, by following only the motion of the ``trunk''
(i.e.\ the red line in figure \ref{fig:branching_decomp}).
We will first work in one dimension.
If the variance of the parent-offspring distance is $\sigma^2$ (and the distribution is not too heavy-tailed),
then the motion of this is approximately Brownian, with variance $\sigma^2$ per generation.
Suppose that the new patch is at distance $x$.
The probability density for the first time such a Brownian motion $B_s$ hits $x$ is
\citep[XXX]{feller}
% note $\sigma B_t \deq B_{\sigma^2 t}$ so $\P\{ \sup_{s\le t} \sigma B_s \le x \} = \P\{ \sup_{s \le \sigma^2 t} B_s \le x \}
\begin{align} \label{eqn:hitting_dist}
  \P\{ B_t=x \;\mbox{and} \; B_u<x \;\mbox{for all}\; u<t\} =  \frac{x}{\sigma^3 t^{3/2}\sqrt{2\pi}} \exp\left(-\frac{x^2}{2t\sigma^2}\right) .
\end{align}
The chance that the descendants of a given migrant arrives at the new patch at time $t$
is now just equation~\eqref{eqn:hitting_dist} multiplied by the chance that 
the migrant family is still alive at time $t$.
For convenience, we define $\alpha = - \log(1-s_m) \approx s_m$,
so that $1-p_e(t) \simeq (1/\mu_K) \exp(-\alpha t)$,
and define $\mu_K$ to be the limiting mean family size conditioned on survival.
By integrating over $t$, we get the chance that descendants of the migrant ever arrive at the new patch is
\begin{align} \label{eqn:estab_prob}
  & \int_0^\infty \frac{(1-s_m)^t}{\mu_K} \frac{x}{\sigma^3 t^{3/2}\sqrt{2\pi}} \exp\left(-\frac{x^2}{2t\sigma^2}\right)  dt \\
  & \qquad = \frac{1}{\mu_K} \exp\left( - \frac{ x \sqrt{\alpha}}{\sigma} \right) .
\end{align}
(Note that this is, up to the constant $\mu_K$, the chance that a Brownian motion killed at rate $\alpha$
ever hits $x$ \citep{feller}.
This approximation holds if $x$ is large enough that the approximation $1-p_e(t) \approx (1-s_m)^t/\mu_K$ holds.)

@@

The probability of establishment of a single migrant in a patch distance $x$ away
is equal to the probability that the migrant's family makes it to the patch at all,
multiplied but the chance that it establishes locally, given that it makes it there.
We have computed the first term in equation~\eqref{eqn:estab_prob},
and we make the small assumption that the second term is just the probability of local fixation
of a beneficial allele present in $K$ copies,
i.e.~$\E[1-(1-p_f(s_b))^K] \approx 2 s_b \mu_K/\xi^2$,
where $p_f(s_b) \approx 2s_b/\xi^2$ is the probability of fixation of a single mutant with additive benefit $s_b$.
Therefore, the probability of establishment is
\begin{align}
  F(x) &\approx \frac{2s_b}{\xi^2} \exp\left( - \frac{\sqrt{\alpha} x}{\sigma} \right) ,
\end{align}
which verifies the less intuitive calculation of $\migrate$ above.

@@

>>>>

Now we can add in the spatial motion, by following only the motion of the ``trunk''
(i.e.\ the red line in figure \ref{fig:branching_decomp}).
We will first work in one dimension.
If the variance of the parent-offspring distance is $\sigma^2$ (and the distribution is not too heavy-tailed),
then the motion of this is approximately Brownian, with variance $\sigma^2$ per generation.
Suppose that the new patch is at distance $x$.
The probability density for the first time such a Brownian motion $B_s$ hits $x$ is
\citep[XXX]{feller}
% note $\sigma B_t \deq B_{\sigma^2 t}$ so $\P\{ \sup_{s\le t} \sigma B_s \le x \} = \P\{ \sup_{s \le \sigma^2 t} B_s \le x \}
\begin{align} \label{eqn:hitting_dist}
  \P\{ B_t=x \;\mbox{and} \; B_u<x \;\mbox{for all}\; u<t\} =  \frac{x}{\sigma^3 t^{3/2}\sqrt{2\pi}} \exp\left(-\frac{x^2}{2t\sigma^2}\right) .
\end{align}
The chance that the descendants of a given migrant arrives at the new patch at time $t$
is now just equation~\eqref{eqn:hitting_dist} multiplied by the chance that 
the migrant family is still alive at time $t$.
For convenience, we define $\alpha = - \log(1-s_m) \approx s_m$,
so that $1-p_e(t) \simeq (1/\mu_K) \exp(-\alpha t)$,
and define $\mu_K$ to be the limiting mean family size conditioned on survival.
By integrating over $t$, we get the chance that descendants of the migrant ever arrive at the new patch is
\begin{align} 
  & \int_0^\infty \frac{(1-s_m)^t}{\mu_K} \frac{x}{\sigma^3 t^{3/2}\sqrt{2\pi}} \exp\left(-\frac{x^2}{2t\sigma^2}\right)  dt \\
  & \qquad = \frac{1}{\mu_K} \exp\left( - \frac{ x \sqrt{\alpha}}{\sigma} \right) .\label{eqn:estab_prob}
\end{align}
(Note that this is, up to the constant $\mu_K$, the chance that a Brownian motion killed at rate $\alpha$
ever hits $x$ \citep{feller}.
This approximation holds if $x$ is large enough that the approximation $1-p_e(t) \approx (1-s_m)^t/\mu_K$ holds.)

@@

The probability of establishment of a single migrant in a patch distance $x$ away
is equal to the probability that the migrant's family makes it to the patch at all,
multiplied but the chance that it establishes locally, given that it makes it there.
We have computed the first term in equation~\eqref{eqn:estab_prob},
and we make the small assumption that the second term is just the probability of local fixation
of a beneficial allele present in $K$ copies,
i.e.~$\E[1-(1-p_f(s_b))^K] \approx 2 s_b \mu_K/\xi^2$,
where $p_f(s_b) \approx 2s_b/\xi^2$ is the probability of fixation of a single mutant with additive benefit $s_b$.
Therefore, the probability of establishment is
\begin{align} \label{eqn:bp_prob_estab}
  F(x) &\approx \frac{2s_b}{\xi^2} \exp\left( - \frac{\sqrt{\alpha} x}{\sigma} \right) ,
\end{align}
which verifies the less intuitive calculation of $\migrate$ above.

@@

\plr{Notes for clarification.}
We can calculate the rate of ``core'' establishment, i.e.\ the mean number of reproductive events that lead to establishment as defined below, in two ways:
\begin{align}
  (\mbox{rate of establishment}) &= (\mbox{outflux of BPs from old patch}) \times (\mbox{\# estab events per BP}) \\
  &= (\mbox{outflux of BPs from old patch}) \times (\mbox{prob of estab of a distant BP}) \\
  & \qquad \qquad \times (\mbox{\# estab events per successful BP}) \\
  &= (\mbox{subcritical occupation density}) \times \rho \times (\mbox{\# offspring per subcritical indiv}) \\
  & \qquad \qquad \times (\mbox{prob of estab in the patch}) .
\end{align}
The probability of establishment of a branching process is in equation~\eqref{eqn:bp_prob_estab};
the subcritical occupation density is in equation~\eqref{eqn:eqfreq}, up to a constant $C$ depending on geometry;
the number of (extra) offspring per subcritical individual is $s_B+s_m$;
and the probability of establishment in the patch is $2s_B/\xi^2$.
The ``outflux of branching processes'' we have not computed;
but we know that the probability a branching process arrives at the patch is given by equation~\eqref{eqn:estab_prob};
if it does it is present in $\mu_K$ copies, each of which leave a total of $1/s_m$ offspring; 
so equating ($\rho$ times equation~\eqref{eqn:eqfreq}) and (outflux times $\mu_K/s_m$ times equation~\eqref{eqn:estab_prob})
we have that the effective outflux (at the edge of the already occupied patch!) is $C \rho s_m$,
rather than $C\rho$ which we might have guessed.
The number of establishment events per BP is as follows, in averages: 
with prob $1/\mu_K \exp(-\sqrt{\alpha} x/ \sigma)$ a total of $\mu_K$ indivs arrive at the new patch; 
each produce $1/s_m$ total core offspring;
each of these produce $s_B+s_m$ extra offspring which each have chance $2s_B/\xi^2$ of establishing.
This squares with the previous argument.
\textbf{Now} note that what we \textbf{actually} want to compute is not this rate, but rather 
\begin{align}
  \migrate &= (\mbox{outflux of BPs from old patch}) \times (\mbox{prob of estab of a BP}) \\
  &= \frac{\rho C}{s_m} \times \frac{2 s_b}{\xi^2} \exp\left(- \frac{\sqrt{\alpha}x}{\sigma} \right) .
\end{align}
\plr{end of notes}
